Are parental care induced gene expression patterns affect more by internal cues or external stimuli?

### 2019-09-07 docs/characterization_manuscript.tex
looking better

\hypertarget{rock-dove-housing}{%
\subsection{Rock dove housing}\label{rock-dove-housing}}

@@

EdgeR and DESeq2 were used to calculate statistical differnces in gene
expression between each transition. All subsequent analyses were
conducting using variance stabilized data from the DESeq2 model. A
linear discriminant analysis was used to determine the confidence with
which we could predict the parental care stage of a sample based on it's
global gene expression profile. WGCNA was used to identify genes who's
expression was correlated with specific candidate genes of interest.

@@

Thosands of genes were found to be differentially expressed between the
non-parental control group and the nest-building group (Table 1, Fig.
1). Within the time course of parental care, fewer than 1000 genes
change there expression over the course of a few days, with the most
pronounced difference in toltal number of changes occuring in the female
pituitary between incubation days 9 and 17 (Table 1, Fig. 2).

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{The magnitude of gene expression changes between each parental
transition. Using DESeq 2, we identified 4-5K genes are differentially
expressed beteen control birds and their nest buliding conspecifics in
all tissues except the male gonad. 500 - 1000 genes are differentially
expressed in the female pituitary from mid-late incubation as well as in
the female hypothalamus and pituitary from hatch to nestingly care day
5.}
\end{figure}

Even though we have a large sample size of roughly 12 samples per tissue
per sex, the differences between all 8 sampled parental time points are
so small, that we are only able to predict the stage using gene
expression along 40\%, 30\% and 27\% of the time in the pituiary, gonads
and hypothalamus, respectively (Fig 3A). However, if we group the
parental stages into anticipatory (control, nest building), egg care
(lay and incubation days), and nestling care (hatch and nesting days),
we incrase our ability to detect parental stage to 47\%, 43\%, and 37
pituiary, hypothalamus and gonads, respectively (Fig 3B). The

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Linear discriminant analysis shows the strongest ability to
predict parental stage of pitutary samples.}
\end{figure}

Gene expression of prlactin (\emph{PRL})in the pitutiary, but not the
hypothalamus follows the characteristic dip in expression relative to
non-parental controls during early incubation then spikes the day before
hatch and peaks at hatch (Fig 4). Intrestingly, hundreds of genes follow
this same, prolactin-associated pattern of expression in the pitutiary
(Fig. 5).

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Prolactin expression in PIT (but not HYP or GON) follows same
pattern as in blood}
\end{figure}

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{Hundreds of genes show similar pattern in expression as
circulating prolactin}
\end{figure}

@@

>>>>

@@

We used EdgeR determine the greatest contributor to variance in gene
expression in the full dataset. Then, we used EdgeR and DESeq2 were used
to calculate statistical differnces in gene expression between each
transition within a given tissue. All subsequent analyses were
conducting using variance stabilized data from the DESeq2 model. A
linear discriminant analysis was used to determine the confidence with
which we could predict the parental care stage of a sample based on it's
global gene expression profile. WGCNA was used to identify genes who's
expression was correlated with specific candidate genes of interest.

@@

We found that the majority of the variation in the full dataset could be
attributed to tissue, with only a notable sex differnce existing between
the female ovaries and the male testies (Fig. 2).

Thousands of genes were found to be differentially expressed between the
non-parental control group and the nest-building group (Table 1, Fig.
3). Within the time course of parental care, fewer than 1000 genes
change there expression over the course of a few days, with the most
pronounced difference in toltal number of changes occuring in the female
pituitary between incubation days 9 and 17 (Table 1, Fig. 3).

Even though we have a large sample size of roughly 12 samples per tissue
per sex, the differences between all 8 sampled parental time points are
so small, that we are only able to predict the stage using gene
expression along 40\%, 30\% and 27\% of the time in the pituiary, gonads
and hypothalamus, respectively (Fig 4A). However, if we group the
parental stages into anticipatory (control, nest building), egg care
(lay and incubation days), and nestling care (hatch and nesting days),
we incrase our ability to detect parental stage to 47\%, 43\%, and 37
pituiary, hypothalamus and gonads, respectively (Fig 4B).

Gene expression of prlactin (\emph{PRL})in the pitutiary, but not the
hypothalamus follows the characteristic dip in expression relative to
non-parental controls during early incubation then spikes the day before
hatch and peaks at hatch (Fig 5).

Hundreds of genes follow this same, prolactin-associated pattern of
expression in the pitutiary (Fig. 6).

@@

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{MDS plots of all samples.}
\end{figure}

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{The magnitude of gene expression changes between each parental
transition. Using DESeq 2, we identified 4-5K genes are differentially
expressed beteen control birds and their nest buliding conspecifics in
all tissues except the male gonad. 500 - 1000 genes are differentially
expressed in the female pituitary from mid-late incubation as well as in
the female hypothalamus and pituitary from hatch to nestingly care day
5.}
\end{figure}

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Linear discriminant analysis shows the strongest ability to
predict parental stage of pitutary samples.}
\end{figure}

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{Prolactin expression in PIT (but not HYP or GON) follows same
pattern as in blood}
\end{figure}

\begin{figure}
\centering
\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{Hundreds of genes show similar pattern in expression as
circulating prolactin}
\end{figure}
